---
title: "Visualize Data"
format: html
editor: visual
---

### Your Turn 1: The ggplot template and aesthetics

Run the following code chunk to import cirrhosis data.

```{r, warning = FALSE, message = FALSE}
library(tidyverse)
cirrhosis <- read_csv("real_data/cirrhosis.csv") 
```

Let's recreate the histogram of `Cholesterol`.

You will need to:

1.  Pick a data set: e.g. *cirrhosis*
2.  Choose a geom function: e.g. *geom_histogram()*
3.  Map aesthetic properties of the graph to variables of the data frame (in this case, the *Cholesterol* variable should be on the x axis).

Fill in the blanks and run the code to create the plot.

```{r}
ggplot(data = cirrhosis) +
  geom_histogram(mapping = aes(x = Cholesterol))
```

When you run this code, you will most likely have what looks like an error displayed above the graph:

`stat_bin()` using `bins = 30`. Pick better value with `binwidth`.

This is actually just a message from R, telling you that it's a bad idea to use a default bin width.

The `Cholesterol` variable holds cholesterol values. What would be an appropriate number of bins or bin width? Try experimenting with various values of either `binwidth` or `bins`.

Fill in the blanks below to recreate the code from the previous chunk, with the *binwidth* argument passed to the *geom_histogram* function.

**Pro Tip**: Often, the most important argument for a function (like the `data` argument for `ggplot` or the `mapping` argument for `geom` functions) is assumed to be first, and you can delete the label.

So, `ggplot(data = cirrhosis)` is the same as `ggplot(cirrhosis)`!

```{r}
# One example, with narrow bin width (each bin is 20 units wide)
ggplot(data = cirrhosis) +
  geom_histogram(mapping = aes(x = Cholesterol), binwidth = 20)
```

```{r}
# A second example, with wide bins (just give me 4 bins)
ggplot(data = cirrhosis) +
  geom_histogram(mapping = aes(x = Cholesterol), bins = 4)
```
### STOP HERE!

Let's next add some color!

Copy and paste the previous code chunk and:

* Set `bins` to 30 (just so we're all on the same page)
* After your first aesthetic mapping, for `x`, add a comma.
* Hit enter, to get a new line (it just makes it look nice)
* Add a second aesthetic mapping after your mapping for `x`: 
  - It will be a mapping for `fill` 
  - It will be set to use the `Sex` column. 
* Makes sure all aesthetics are within the `aes()` function.

```{r}
ggplot(data = cirrhosis) +
  geom_histogram(mapping = aes(x = Cholesterol,
                               fill = Sex), bins = 30)
```

### Your Turn: Geom functions

Run the following code chunk, which should look familiar to you in a few ways!

```{r}
ggplot(data = cirrhosis) +
  geom_histogram(mapping = aes(x = N_Days),
                 binwidth = 365)
```

Now try to figure out how you would modify the code above so it draws a frequency polygon instead of histogram.

```{r}
ggplot(data = cirrhosis) +
  geom_freqpoly(mapping = aes(x = N_Days),
                binwidth = 365)
```

Modify the previous code chunk so that the frequency polygon's line color is blue, regardless of the data:

* Inside the `geom` function (NOT inside the `aes`, find your last argument (probably `binwidth`), and after that argument, put a comma.
* Put the argument name: `color`
* Set it equal to the color you want. Color names go in quotes!

```{r}
ggplot(data = cirrhosis) +
  geom_freqpoly(mapping = aes(x = N_Days),
                binwidth = 365,
                color = "blue")
```

Now modify the frequency polygon so that you have two lines: one for men, and one for women.  How would you do this?  Would it go inside or outside the `aes` function?

```{r}
ggplot(data = cirrhosis) +
  geom_freqpoly(mapping = aes(x = N_Days,
                              color = Sex),
                binwidth = 365)
```

What do you think the following code will do? Try to predict what you'll see.

Then run it.

```{r}
ggplot(data = cirrhosis) +
  geom_histogram(mapping = aes(x = N_Days),
                 binwidth = 365) +
  geom_freqpoly(mapping = aes(x = N_Days,
                              color = Sex),
                binwidth = 365)
```